---
title: "Temperature effects on fat biking, part two"
author: "Christopher Swingley"
date: '2015-12-21'
output:
  pdf_document:
    highlight: default
    includes:
      in_header: rmarkdown_tex_preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    toc: no
  html_document:
    theme: united
    toc: yes
  word_document: default
---

# Introduction

While riding to work this morning I figured out a way to disentangle the effects
of trail quality and physical conditioning (both of which improve over the
season) from temperature, which also tends to increase throughout the season.
As you recall in my previous post, I found that days into the season (winter day
of year) and minimum temperature were both negatively related with fat bike
energy consumption.  But because those variables are related, we can’t make
statements about them individually.

But what if we look at pairs of trips that are within two days of each other and
look at the difference in temperature between those trips and the difference in
energy consumption?  We’ll only pair trips going the same direction (to or from
work), and we’ll restrict the pairings to two days or less.  That eliminates
seasonality from the data because we’re always comparing two trips from the same
few days.

# Data

For this analysis, I’m using SQL to filter the data because I’m better at window
functions and filtering in SQL than R.  Here’s the code to grab the data from
the database.  (The CSV is on my [GitHub
repo](https://github.com/cswingle/fat_biking) for this analysis).  The trick
here is to categorize trips as being to work (“north”) or from work (“south”)
and then include this field in the partition statement of the window function so
I’m only getting the next trip that matches direction.

```{r echo=TRUE, eval=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(scales)

exercise_db <- src_postgres(host="example.com", dbname="exercise_data")

diffs <- tbl(exercise_db,
             build_sql(
   "WITH all_to_work AS (
        SELECT *,
            CASE WHEN extract(hour from start_time) < 11 THEN 'north' ELSE 'south' END AS direction
        FROM track_stats
        WHERE type = 'Fat Biking'
            AND miles between 4 and 4.3
    ), with_next AS (
        SELECT track_id, start_time, direction, kcal, miles, min_temp,
            lead(direction) OVER w AS next_direction,
            lead(start_time) OVER w AS next_start_time,
            lead(kcal) OVER w AS next_kcal,
            lead(miles) OVER w AS next_miles,
            lead(min_temp) OVER w AS next_min_temp
        FROM all_to_work
        WINDOW w AS (PARTITION BY direction ORDER BY start_time)
    )
    SELECT start_time, next_start_time, direction,
        min_temp, next_min_temp,
        kcal / miles AS kcal_per_mile,
        next_kcal / next_miles as next_kcal_per_mile,
        next_min_temp - min_temp AS temp_diff,
        (next_kcal / next_miles) - (kcal / miles) AS kcal_per_mile_diff
    FROM with_next
    WHERE next_start_time - start_time < '60 hours'
    ORDER BY start_time")) %>% collect()

write.csv(diffs, file="fat_biking_trip_diffs.csv", quote=TRUE,
          row.names=FALSE)
kable(head(diffs))
```

```{r echo=FALSE, eval=TRUE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
diffs <- read.csv("fat_biking_trip_diffs.csv", header=TRUE,
                  stringsAsFactors=FALSE)
kable(head(diffs))
```

Out of a total of 123 trips, 70 took place within 2 days of each other.  We
still don’t have a measure of trail quality, so pairs where the trail is smooth
and hard one day and covered with fresh snow the next won’t be particularly good
data points.

Let’s look at a plot of the data.

```{r echo=TRUE, eval=TRUE, warning=FALSE}
s = ggplot(data=diffs,
           aes(x=temp_diff, y=kcal_per_mile_diff)) +
    geom_point() +
    geom_smooth(method="lm", se=FALSE) +
    scale_x_continuous(name="Temperature difference between paired trips (degrees F)",
                       breaks=pretty_breaks(n=10)) +
    scale_y_continuous(name="Energy consumption difference (kcal / mile)",
                       breaks=pretty_breaks(n=10)) +
    theme_bw() +
    ggtitle("Paired fat bike trips to and from work within 2 days of each other")

print(s)
```

What this shows is that when the temperature difference between two paired trips
is negative (the second trip is colder than the first), additional energy is
required for the second (colder) trip.  This matches the pattern we saw in my
last post where minimum temperature and winter day of year were negatively
associated with energy consumption.  But because we’ve used differences to
remove seasonal effects, we can actually determine how large of an effect
temperature has.

There are quite a few outliers here.  Those that are in the region with very
little difference in temperature are likey due to snowfall changing the trail
conditions from one trip to the next.

I’m not sure why there is so much scatter among the points on the left side of
the graph.  I don’t see any particular pattern among those points that might
explain the higher than normal variation (see the table below), and we don’t see
the same variation in the points with a large positive difference in
temperature, so I think this is normal variation in the data not explained by
temperature.

```{r echo=TRUE, eval=TRUE, warning=FALSE}
diffs %>% filter(temp_diff < -20) %>% kable()
```

# Results

Here’s the linear regression results for this data.

```{r echo=TRUE, eval=TRUE, warning=FALSE}
summary(lm(data=diffs, kcal_per_mile_diff ~ temp_diff))
```

The model and coefficient are both highly signficant, and as we might expect,
the intercept in the model is not significantly different from zero (if there
wasn’t a difference in temperature between two trips there shouldn’t be a
difference in energy consumption either, on average).  Temperature alone
explains 30% of the variation in energy consumption, and the coefficient tells
us the scale of the effect: each degree drop in temperature results in an
increase in energy consumption of 0.78 kcalories per mile.  So for a 4 mile
commute like mine, the difference between a trip at 10°F vs −20°F is an
additional 93 kilocalories (30 × 0.7778 × 4 = 93.34) on the colder trip.  That
might not sound like much in the context of the calories in food (93
kilocalories is about the energy in a large orange or a light beer), but my
average energy consumption across all fat bike trips to and from work is 377
kcalories so 93 represents a large portion of the total.


<!-- vim: set ft=rmd fenc=utf-8 tw=80 ts=4 sw=4 sts=4 :-->
